---
title: "Schizophrenia GBD Analysis Results"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: yes
    df_print: paged
---

```{r rmd-setting, echo=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 150
)
```

This report mirrors the breast cancer analysis structure, adapted to the schizophrenia extract. The schizophrenia cause table contains incidence, prevalence, YLDs, and DALYs for China and the United States, with female and male strata. The source age structure uses `70+ years` as the top age bin. The report focuses on the available cause-burden measures in this extract.

```{r setup}
library(dplyr)
library(tidyr)
library(readr)
library(knitr)

options(scipen = 999)

input_file <- knitr::current_input()
candidate_inputs <- unique(c(
  input_file,
  file.path(getwd(), input_file),
  file.path("schizo_GBD", basename(input_file))
))
candidate_inputs <- candidate_inputs[!is.na(candidate_inputs) & file.exists(candidate_inputs)]

root_dir <- if (length(candidate_inputs) > 0) {
  dirname(normalizePath(candidate_inputs[[1]], winslash = "/", mustWork = TRUE))
} else {
  normalizePath(".", winslash = "/", mustWork = TRUE)
}
if (!dir.exists(file.path(root_dir, "derived")) && dir.exists(file.path(root_dir, "schizo_GBD", "derived"))) {
  root_dir <- file.path(root_dir, "schizo_GBD")
}
derived_dir <- file.path(root_dir, "derived")
fig_dir <- file.path(root_dir, "figures")

locations <- c("China", "United States of America")
sexes <- c("Female", "Male")
main_measures <- c("Incidence", "Prevalence", "DALYs")
all_measures <- c("Incidence", "Prevalence", "YLDs", "DALYs")

fmt_int <- function(x) {
  format(round(as.numeric(x)), big.mark = ",", scientific = FALSE, trim = TRUE)
}

fmt_num <- function(x, digits = 1) {
  x_num <- as.numeric(x)
  ifelse(
    is.na(x_num),
    "NA",
    formatC(x_num, digits = digits, format = "f", big.mark = ",")
  )
}

fmt_pct <- function(x, digits = 1) {
  paste0(fmt_num(x, digits), "%")
}

fmt_signed_int <- function(x) {
  x_num <- as.numeric(x)
  paste0(ifelse(!is.na(x_num) & x_num > 0, "+", ""), fmt_int(x_num))
}

fmt_signed_pct <- function(x, digits = 1) {
  x_num <- as.numeric(x)
  paste0(ifelse(!is.na(x_num) & x_num > 0, "+", ""), fmt_pct(x_num, digits))
}

fmt_ratio <- function(x, digits = 2) {
  formatC(as.numeric(x), digits = digits, format = "f")
}

fmt_ratio_range <- function(x, digits = 2) {
  x_num <- as.numeric(x)
  x_num <- x_num[!is.na(x_num)]
  if (length(x_num) == 0) {
    return("NA")
  }
  paste0(fmt_ratio(min(x_num), digits), "-", fmt_ratio(max(x_num), digits))
}

fmt_pct_range <- function(x, digits = 2) {
  x_num <- as.numeric(x)
  x_num <- x_num[!is.na(x_num)]
  if (length(x_num) == 0) {
    return("NA")
  }
  paste0(fmt_signed_pct(min(x_num), digits), " to ", fmt_signed_pct(max(x_num), digits))
}

fmt_count_range <- function(x) {
  x_num <- as.numeric(x)
  x_num <- x_num[!is.na(x_num)]
  if (length(x_num) == 0) {
    return("NA")
  }
  if (min(x_num) == max(x_num)) {
    return(fmt_int(min(x_num)))
  }
  paste0(fmt_int(min(x_num)), "-", fmt_int(max(x_num)))
}

format_joinpoints <- function(x, empty = "None") {
  value <- trimws(as.character(x))
  value[is.na(value)] <- ""
  ifelse(value == "", empty, gsub(",", ", ", value))
}

emit_burden_highlights <- function(data) {
  cat("Selected endpoint highlights:\n\n")
  for (sx in sexes) {
    for (measure in main_measures) {
      rows <- data %>%
        filter(sex_name == sx, measure_name == measure)
      china <- rows %>% filter(location_name == "China") %>% slice_head(n = 1)
      us <- rows %>% filter(location_name == "United States of America") %>% slice_head(n = 1)
      if (nrow(china) == 0 || nrow(us) == 0) {
        next
      }
      cat(
        "- **", sx, " ", measure, ".** China reached ", fmt_int(china$count_2023),
        " in 2023 (", fmt_signed_pct(china$count_change_pct, 1), " since 1990; ASR ",
        fmt_num(china$asr_2023, 2), " per 100,000). The United States reached ",
        fmt_int(us$count_2023), " (", fmt_signed_pct(us$count_change_pct, 1),
        "; ASR ", fmt_num(us$asr_2023, 2), " per 100,000).\n",
        sep = ""
      )
    }
  }
  cat("\n")
}

emit_comparison_highlights <- function(data) {
  data_main <- data %>% filter(measure_name %in% main_measures)
  country_asr <- data_main %>%
    filter(comparison_type == "China vs United States", metric_name == "Age-standardized rate")
  country_counts <- data_main %>%
    filter(comparison_type == "China vs United States", metric_name == "All-age number")
  sex_asr <- data_main %>%
    filter(comparison_type == "Male vs Female", metric_name == "Age-standardized rate")
  sex_counts <- data_main %>%
    filter(comparison_type == "Male vs Female", metric_name == "All-age number")
  largest_shifts <- data_main %>%
    mutate(abs_shift = abs(ratio_change_pct)) %>%
    arrange(desc(abs_shift)) %>%
    slice_head(n = 3) %>%
    transmute(
      label = paste0(
        comparison_type, ", ", sex_name, " ", measure_name, " ",
        metric_name, " (", fmt_signed_pct(ratio_change_pct, 1), ")"
      )
    )

  cat("Concise contrast summary:\n\n")
  cat(
    "- **Country rate contrast.** In 2023, China/United States age-standardized rate ratios for the main measures ranged from ",
    fmt_ratio_range(country_asr$ratio_2023, 2),
    "; values below 1.00 indicate lower ASRs in China than in the United States.\n",
    sep = ""
  )
  cat(
    "- **Country count contrast.** China's 2023 all-age numbers were ",
    fmt_ratio_range(country_counts$ratio_2023, 2),
    " times the corresponding U.S. counts across the main measures.\n",
    sep = ""
  )
  cat(
    "- **Sex contrast.** Male/female 2023 ratios ranged from ",
    fmt_ratio_range(sex_asr$ratio_2023, 2),
    " for ASRs and ", fmt_ratio_range(sex_counts$ratio_2023, 2),
    " for all-age numbers.\n",
    sep = ""
  )
  cat("- **Largest ratio shifts.** ", paste(largest_shifts$label, collapse = "; "), ".\n\n", sep = "")
}

emit_segmented_highlights <- function(data) {
  segmented_notes <- data %>%
    filter(measure_name %in% main_measures) %>%
    group_by(measure_name) %>%
    summarise(
      panels = n(),
      panels_with_joinpoints = sum(joinpoint_count > 0, na.rm = TRUE),
      aapc_range = fmt_pct_range(aapc, 2),
      eapc_range = fmt_pct_range(single_slope_eapc, 2),
      .groups = "drop"
    ) %>%
    arrange(factor(measure_name, levels = main_measures))

  cat("Segmented-model summary:\n\n")
  for (row_index in seq_len(nrow(segmented_notes))) {
    row <- segmented_notes[row_index, ]
    cat(
      "- **", row$measure_name, ".** AAPC ranged from ", row$aapc_range,
      " across ", row$panels, " location-sex panels; ",
      row$panels_with_joinpoints, " panels selected at least one joinpoint. The comparable single-slope EAPC range was ",
      row$eapc_range, ".\n",
      sep = ""
    )
  }
  cat("\n")
}

emit_decomposition_highlights <- function(data) {
  component_notes <- data %>%
    filter(measure_name %in% main_measures) %>%
    mutate(
      dominant_component = case_when(
        abs(population_growth) >= abs(population_aging) & abs(population_growth) >= abs(rate_change) ~ "population growth",
        abs(population_aging) >= abs(population_growth) & abs(population_aging) >= abs(rate_change) ~ "population aging",
        TRUE ~ "rate change"
      ),
      dominant_value = case_when(
        dominant_component == "population growth" ~ population_growth,
        dominant_component == "population aging" ~ population_aging,
        TRUE ~ rate_change
      ),
      dominant_share = case_when(
        dominant_component == "population growth" ~ population_growth_pct_of_total,
        dominant_component == "population aging" ~ population_aging_pct_of_total,
        TRUE ~ rate_change_pct_of_total
      )
    ) %>%
    group_by(measure_name) %>%
    slice_max(abs(total_change), n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(factor(measure_name, levels = main_measures))

  cat("Endpoint decomposition highlights:\n\n")
  for (row_index in seq_len(nrow(component_notes))) {
    row <- component_notes[row_index, ]
    cat(
      "- **", row$measure_name, ".** The largest adult count change was in ",
      row$location_name, " ", row$sex_name, " (", fmt_signed_int(row$total_change),
      "). The dominant component was ", row$dominant_component, " (",
      fmt_signed_int(row$dominant_value), "; ", fmt_signed_pct(row$dominant_share, 1),
      " of total change).\n",
      sep = ""
    )
  }
  cat("\n")
}

emit_annual_chain_highlights <- function(data) {
  data_main <- data %>% filter(measure_name %in% main_measures)
  max_total_difference <- max(abs(data_main$total_change_difference), na.rm = TRUE)
  path_shift <- data_main %>%
    mutate(
      component_shift =
        abs(annual_chain_population_growth - endpoint_population_growth) +
        abs(annual_chain_population_aging - endpoint_population_aging) +
        abs(annual_chain_rate_change - endpoint_rate_change)
    ) %>%
    slice_max(component_shift, n = 1, with_ties = FALSE)

  cat("Annual-chain decomposition checks:\n\n")
  cat(
    "- **Total-change agreement.** Annual-chain totals matched endpoint totals within ",
    fmt_num(max_total_difference, 6), " counts across the main panels.\n",
    sep = ""
  )
  cat(
    "- **Path dependence.** Component allocation shifted most for ",
    path_shift$location_name, " ", path_shift$sex_name, " ", path_shift$measure_name,
    ": endpoint rate change was ", fmt_signed_int(path_shift$endpoint_rate_change),
    " versus ", fmt_signed_int(path_shift$annual_chain_rate_change),
    " under annual chaining.\n\n",
    sep = ""
  )
}

emit_apc_coverage_highlights <- function(cells, summary_data) {
  coverage <- cells %>%
    group_by(location_name, sex_name, measure_name) %>%
    summarise(
      age_groups = n_distinct(age_name),
      period_blocks = n_distinct(period_label),
      cohort_groups = n_distinct(cohort_label),
      age_period_cells = n(),
      .groups = "drop"
    ) %>%
    left_join(
      summary_data %>%
        select(location_name, sex_name, measure_name, model_year_start, model_year_end),
      by = c("location_name", "sex_name", "measure_name")
    ) %>%
    filter(measure_name %in% main_measures)

  year_span <- coverage %>%
    summarise(
      start_year = min(model_year_start, na.rm = TRUE),
      end_year = max(model_year_end, na.rm = TRUE)
    )

  cat("APC coverage summary:\n\n")
  cat(
    "- Main-measure APC panels covered ", year_span$start_year, "-", year_span$end_year,
    " with ", fmt_count_range(coverage$age_groups), " age groups, ",
    fmt_count_range(coverage$period_blocks), " period blocks, ",
    fmt_count_range(coverage$cohort_groups), " cohort groups, and ",
    fmt_count_range(coverage$age_period_cells), " age-period cells per panel.\n",
    sep = ""
  )
  cat(
    "- Reference ages, periods, and cohorts remain available in the APC summary CSV, but the full panel matrix is omitted from the report body for readability.\n\n"
  )
}

emit_apc_highlights <- function(summary_data, local_drift_data) {
  net_notes <- summary_data %>%
    filter(measure_name %in% main_measures) %>%
    mutate(abs_net_drift = abs(net_drift)) %>%
    arrange(desc(abs_net_drift)) %>%
    slice_head(n = 4) %>%
    transmute(
      label = paste0(location_name, " ", sex_name, " ", measure_name, " (", fmt_signed_pct(net_drift, 2), ")")
    )

  local_notes <- local_drift_data %>%
    filter(measure_name %in% main_measures) %>%
    mutate(abs_local_drift = abs(local_drift)) %>%
    arrange(desc(abs_local_drift)) %>%
    slice_head(n = 4) %>%
    transmute(
      label = paste0(
        location_name, " ", sex_name, " ", measure_name, " at ", age_name,
        " (", fmt_signed_pct(local_drift, 2), ")"
      )
    )

  cat("APC comparative highlights:\n\n")
  cat("- **Net drift.** Largest absolute net-drift estimates: ", paste(net_notes$label, collapse = "; "), ".\n", sep = "")
  cat("- **Local drift.** Strongest local-drift estimates: ", paste(local_notes$label, collapse = "; "), ".\n\n", sep = "")
}

read_derived <- function(filename) {
  read_csv(file.path(derived_dir, filename), show_col_types = FALSE)
}

include_figure <- function(filename) {
  path <- file.path(fig_dir, filename)
  if (file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    cat("Missing figure:", path)
  }
}
```

```{r load-data}
burden_audit <- read_derived("burden_data_audit_1990_2023.csv")
burden_summary <- read_derived("burden_summary_1990_2023.csv")
comparative_metrics <- read_derived("comparative_metrics_1990_2023.csv")
segmented_summary <- read_derived("segmented_summary_1990_2023.csv")
decomposition <- read_derived("decomposition_1990_2023.csv")
decomposition_interval <- read_derived("decomposition_interval_1990_2023.csv")
annual_chain_final <- read_derived("annual_chain_final_summary.csv")
annual_chain_vs_endpoint <- read_derived("annual_chain_vs_endpoint_1990_2023.csv")
age_specific <- read_derived("age_specific_comparison_1990_2023.csv")
age_difference <- read_derived("age_specific_china_us_difference_1990_2023.csv")
apc_summary <- read_derived("apc_summary_1990_2023.csv")
apc_age_curve <- read_derived("apc_age_curve_1990_2023.csv")
apc_local_drift <- read_derived("apc_local_drift_1990_2023.csv")
apc_period_rr <- read_derived("apc_period_rr_1990_2023.csv")
apc_cohort_rr <- read_derived("apc_cohort_rr_1990_2023.csv")
apc_cells <- read_derived("apc_input_cells_1990_2023.csv")
```


# Overall Burden Trends

Counts are all-age values and rates are age-standardized rates per 100,000. The `Percent basis` column identifies the age basis of the GBD `Percent` metric used for each measure; in this extract, incidence and prevalence use age-standardized percent rows when available, whereas DALYs and YLDs use all-age percent rows. DALY and YLD numbers and rates are numerically identical here because the extract has no YLL component for schizophrenia; their percent values can still differ because the GBD percent denominator is measure-specific.

```{r burden-summary-notes, results='asis'}
emit_burden_highlights(burden_summary)
```

```{r count-trend-figure, out.width='100%'}
include_figure("01_trend_counts.png")
```

```{r percent-trend-figure, out.width='100%'}
include_figure("01b_trend_percent.png")
```

```{r asr-trend-figure, out.width='100%'}
include_figure("02_trend_asr.png")
```

# Country And Sex Contrasts

Ratios compare like-for-like point estimates within each metric. The `Measure-specific GBD percent` rows should be read as ratios of the GBD `Percent` metric for the same measure and age basis, not as a single common "share of all-cause burden" denominator across incidence, prevalence, YLDs, and DALYs.

```{r comparison-notes, results='asis'}
emit_comparison_highlights(comparative_metrics)
```

# Segmented Trends

Age-standardized rates are fitted with a compact BIC-selected segmented log-linear model, allowing up to two joinpoints per location-sex-measure series. The single-slope EAPC is retained alongside the segmented AAPC for continuity with common GBD trend reporting.

```{r segmented-notes, results='asis'}
emit_segmented_highlights(segmented_summary)
```

```{r segmented-figure, out.width='100%'}
include_figure("03_segmented_trends.png")
```

# Age-Specific Patterns

The age-specific panels retain the schizophrenia-specific top bin (`70+ years`). The heatmap difference panels show China relative to the United States within sex, measure, age, and year. Percent panels use the GBD `Percent` metric for each measure.

```{r age-number-figure, out.width='100%'}
include_figure("05b_age_specific_numbers_1990_vs_2023.png")
```

```{r age-number-difference-figure, out.width='100%'}
include_figure("05c_age_specific_number_difference.png")
```

```{r age-percent-figure, out.width='100%'}
include_figure("05d_age_specific_percent_1990_vs_2023.png")
```

```{r age-percent-difference-figure, out.width='100%'}
include_figure("05e_age_specific_percent_difference.png")
```

```{r age-rate-figure, out.width='100%'}
include_figure("04_age_specific_rate_1990_vs_2023.png")
```

```{r age-rate-difference-figure, out.width='100%'}
include_figure("05_age_specific_rate_difference.png")
```

## Full Age-Year Heatmaps

```{r age-number-heatmap, out.width='100%'}
include_figure("05f_age_specific_number_heatmaps_1990_2023.png")
```

```{r age-percent-heatmap, out.width='100%'}
include_figure("05g_age_specific_percent_heatmaps_1990_2023.png")
```

```{r age-rate-heatmap, out.width='100%'}
include_figure("05h_age_specific_rate_heatmaps_1990_2023.png")
```

# Decomposition

The decomposition uses adult age-specific numbers and rates from 15-19 years through 70+ years. It decomposes 1990-2023 adult count change into adult population growth, adult population aging, and age-specific rate change using the average replacement approach. Population denominators are calculated from GBD count/rate pairs.

```{r decomposition-notes, results='asis'}
emit_decomposition_highlights(decomposition)
```

```{r decomposition-figure, out.width='100%'}
include_figure("06_decomposition.png")
```

# Annual Chained Decomposition

The annual chained decomposition repeats the same population-growth, population-aging, and rate-change replacement decomposition for adjacent years, then accumulates components from 1990. As in the breast cancer analysis, component allocations differ from the direct 1990-2023 endpoint decomposition because the decomposition is path-dependent; total observed change should still agree apart from numerical tolerance.

```{r annual-chain-notes, results='asis'}
emit_annual_chain_highlights(annual_chain_final)
```

```{r annual-chain-figure, out.width='100%'}
include_figure("08_annual_chained_decomposition.png")
```

```{r interval-chain-figure, out.width='100%'}
include_figure("09_interval_chained_decomposition.png")
```

# Age-Period-Cohort Analysis

This section follows the breast cancer report's APC layout. Annual age-specific schizophrenia observations from 1990-2023 are collapsed into recent-aligned five-year period blocks (`1990-1993`, `1994-1998`, ..., `2019-2023`) and summarized through net drift, age curves, local drift, period relative risks, and cohort relative risks.

## Surface Coverage

The APC model coverage is summarized below without the full reference-group matrix.

```{r apc-coverage-notes, results='asis'}
emit_apc_coverage_highlights(apc_cells, apc_summary)
```

## Incidence and DALY Age-Period Surfaces: China and United States

The age-period rate surfaces are shown with China and the United States side by side within each sex and measure panel.

```{r apc-figure, out.width='100%'}
include_figure("07_apc_rate_heatmaps.png")
```

## Marginal APC Summaries: China and United States

The marginal panels show fitted age curves, local drift, period relative risks, and cohort relative risks by measure and sex.

```{r apc-marginal-figure, out.width='100%'}
include_figure("11_apc_marginal_effects.png")
```

## Cohort Relative-Risk Patterns: China and United States

The cohort panels compare China and the United States within each sex and measure.

```{r apc-cohort-figure, out.width='100%'}
include_figure("12_apc_cohort_rr.png")
```

## Comparative Summary

```{r apc-summary-notes, results='asis'}
emit_apc_highlights(apc_summary, apc_local_drift)
```
